---
title: "Making HTML Tables with gtsummary"
output:
  html_document: default
---

Using the [*gtsummary* package](https://education.rstudio.com/blog/2020/07/gtsummary/) to make various types of tables in r markdown.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning = F, message = F}
library(dplyr)
library(gtsummary)
library(palmerpenguins)
library(psych)
library(tidyverse)
# data(package = 'palmerpenguins')
```

## Summarize descriptive statistics

Throughout the post we will use an example dataset of 200 subjects treated with either Drug A or Drug B, with a mix of categorical, dichotomous, and continuous demographic and response data. The dataset has label attributes (using the labelled package) for column names.

This markdown will use the [palmer penguins dataset](https://allisonhorst.github.io/palmerpenguins/)

```{r}
sm_penguins <- penguins %>% select(species, island, sex, flipper_length_mm, body_mass_g)

head(sm_penguins)
```
In one line of code we can summarize the overall demographics of the dataset!

Notice some nice default behaviors:
Detects variable types of input data and calculates descriptive statistics
Variables coded as 0/1, TRUE/FALSE, and Yes/No are presented dichotomously
Recognizes NA values as “missing” and lists them as unknown
Label attributes automatically printed
Variable levels indented and footnotes added

```{r}
tbl_summary(sm_penguins)
```


Start customizing by adding arguments and functions

Next you can start to customize the table by using arguments of the tbl_summary() function, as well as pipe the table through additional gtsummary functions to add more information, like p-value to compare across groups and overall demographic column.

```{r}
sm_penguins %>%
  tbl_summary(by = island) %>%
  add_p() %>%
  add_overall() %>%
  bold_labels()
```


Customize further using formula syntax and tidy selectors

Most arguments to tbl_summary() and tbl_regression() require formula syntax:

select variables ~ specify what you want to do

To select, use quoted or unquoted variables, or minus sign to negate (e.g. age or "age" to select, -age to deselect)
Or use any {tidyselect} functions, e.g. contains("stage") ~ ..., including type selectors
To specify what you want to do, some arguments use {glue} syntax where whatever is in the curly brackets gets evaluated and passed directly into the string. e.g statistic = ... ~ "{mean} ({sd})"

```{r, warning = F, message = F}
sm_penguins %>%
  tbl_summary(
    by = sex,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} / {N} ({p}%)"
    ),
    label = species ~ "Penguin Species"
  ) %>%
  add_p(
    test = all_continuous() ~ "t.test",
    pvalue_fun = function(x) style_pvalue(x, digits = 2)
  )
```

## Summarize regression models 

### Logistic
First, create a logistic regression model to use in examples.

```{r}
m1 <- glm(sex ~ body_mass_g + species,
  data = penguins,
  family = binomial
)
```
tbl_regression() accepts regression model object as input. Uses {broom} in the background, outputs table with nice defaults:


Reference groups added to the table
Sensible default number rounding and formatting
Label attributes printed
Common model types detected and appropriate header added with footnote

```{r}
tbl_regression(m1, exponentiate = TRUE)
```

### Linear
First, create a linear regression model to use in examples.

```{r}
m2 <- lm(body_mass_g ~ sex + species + island,
  data = penguins
)
```
tbl_regression() accepts regression model object as input. Uses {broom} in the background, outputs table with nice defaults:


Reference groups added to the table
Sensible default number rounding and formatting
Label attributes printed
Common model types detected and appropriate header added with footnote

```{r}
tbl_regression(m2)
```

## Join two or more tables 
Oftentimes we must present results for multiple outcomes of interest, and there are many other reasons you might want to join two summary tables together. We’ve got you covered!

In this example we can use tbl_merge() to merge two gtsummary objects side-by-side. There is also a tbl_stack() function to place tables on top of each other.

```{r}
m3 <- lm(flipper_length_mm ~ sex + species + island,
  data = penguins
)

tbl_reg_1 <- tbl_regression(m2)

tbl_reg_2 <- tbl_regression(m3)

tbl_merge(
    tbls = list(tbl_reg_1, tbl_reg_2), 
    tab_spanner = c("**Body Mass**", "**Flipper Length**") 
  ) 
```

## Report results inline 
Tables are important, but we often need to report results in-line in a report. Any statistic reported in a gtsummary table can be extracted and reported in-line in a R Markdown document with the inline_text() function.

`inline_text(tbl_reg_1, variable = species, level = "Gentoo")` 

**produces:**

`r inline_text(tbl_reg_1, variable = species, level = "Gentoo")`